\chapter{Related work}
\label{chap:related}
\section{Heap-shape information}
Separation logic~\cite{Reynolds:2002:SLL:645683.664578} is an extension of Hoare logic for reasoning about shared data structures. Its key feature is \emph{locality} where proofs or specifications of a subprogram in separation logic only mention the memory portion that is actually used in the subprogram. Applications include program analyses such as compositional shape analysis ~\cite{Calcagno:2009:CSA:1594834.1480917} which allows each procedure in a program to be analysed separately. The techniques developed in compositional shape analysis are in turn used e.g. in ~\cite{fbinfer}.

AProVE~\cite{automated-termination-proofs-for-java-bytecode-with-cyclic-data}, Julia~\cite{DBLP:journals/scp/ScapinS14} or COSTA~\cite{GenaimZ13} all developed tools that deal with heap-shape information, whether using term rewriting systems in AProVE, abstract interpretation combined with constraint graphs in Julia or abstract interpretation with propositional formulae in COSTA.

Additionally, graph grammar methods use hyperedge replacement grammars~\cite{Heinen2015157} to obtain a finite hypergraph representation of data structures. This yields a finite state space that can then be further analysed with e.g. model checking tools.

\section{Points-to analysis}
Points-to analysis, also known as pointer analysis, is a static analysis that calculates the set of references that a reference can possibly take at runtime. Algorithms in weakly typed languages such as Steensgaard's method~\cite{Steensgaard:1996:PAA:237721.237727} cannot use type information as weakly typed languages support arbitrary casts. However in strongly typed, object-oriented languages such as C++ or Java algorithms such as rapid type analysis~\cite{Bacon:1996:FSA:236338.236371} or variable type analysis~\cite{Sundaresan:2000:PVM:354222.353189} do utilize and store type information, which can then be used for e.g. call devirtualization. These analyses either ignore fields completely in case of a \emph{field-insensitive} points-to analysis or do not allow enumerating all fields at which some object or type occur as is the case for a \emph{field-sensitive} points-to analysis.